---
title: "Basics of timepls R package"
author: "scworland@usgs.gov"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: paper
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=F}
knitr::opts_knit$set(warnings=F,message=F)

library(dplyr)
library(ggplot2)
library(tidyr)
library(timepls)
```

## Introduction 

This package was created to model time series using (A) multiple covariates, and (B) when the *exact* number of optimal lags is unknown. For example, you are hoping to predict daily streamflow using precipitation, minimum temperature, maximum temperature, and mean temperature. The time series might look something like this:

```{r,echo=F,fig.align='center',fig.width=5,fig.height=5}
d <- climate_data
dplot <- select(d,date,x1_precip=p,x2_tmin=tmin,x3_tmax=tmax,
                x4_tmean=tmean,y_streamflow=cfs) %>%
  gather(key,value,-date)

ggplot(dplot, aes(date, value)) + 
  geom_line(color="dodgerblue") + 
  facet_wrap(~key, scales="free_y", ncol=1) +
  theme_bw() +
  scale_x_date(date_breaks="3 years",date_labels = "%Y") +
  theme(panel.background = element_rect(fill = "white", colour = "black"),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.grid = element_blank())
```

In this situation, it is not clear what variables are the most important and what lags should be used. Hydrologic theory tells use that streamflow is influenced by antecedent climate conditions, and if we know something about the system, we could probably make a few reasonable assumptions and be ok using a regression model with specific inputs that we chose apriori. But what if we are trying to predict streamflow for multiple sites with very different average climates? What if some basins are influenced by snowmelt? How do we account for the effect of different infiltration capacities of soils? What if the relationships change through time? This package is being developed to be used in these situations. 

## Method





